### 2020-06-04 writeup/transfer.tex
sec 3 fix

\textit{The isotropic model.}
	Consider two tasks with isotropic covariances $\Sigma_1 = \Sigma_2 = \id$.
	Each task has sample size $n_i = \rho_i \cdot p$, for $i = 1, 2$.
	%And $X_1\in\real^{n_1\times p}, X_2\in\real^{n_2\times p}$ denote the covariates of the two tasks, respectively.
	We assume that for the target task, $\beta_2$ has i.i.d. entries with mean zero and variance $\kappa^2$.
	For the source task, $\beta_1 $ equals $\beta_2$ plus i.i.d. entries with mean $0$ and variance $d^2$.
	The labels are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\e_i$ consists of i.i.d. entries with mean zero and variance $\sigma_i^2$, for $i=1,2$.
%	We assume that all the random variables have subexponential decay, while keeping in mind that our results can be applied under weaker moments assumptions as shown in Appendix \ref{sec_maintools}.

In the isotropic model, we show that as we increase the distance between $\beta_1$ and $\beta_2$, there is a transition from positive transfer to negative transfer in MTL.
%We measure model dissimilarity as $\norm{\beta_1 - \beta_2}^2$, which is the distance between source and target in the isotropic model.
%Figure \ref{fig_model_shift} provides a simulation when $p = 200$.
%{The rest of parameter settings can be found in Appendix \ref{app_synthetic}.}
%Our result below will provide an explanation to this phenomenon.
%Based on Theorem \ref{thm_model_shift}, we derive the transition threshold in the following proposition.
We introduce the following notations.
{\small\begin{align*}
	\Psi(\beta_1, \beta_2) = {\ex{\bignorm{\beta_1 - \beta_2}^2}} / {\sigma^2},  \quad \Phi(\rho_1, \rho_2) = \frac{(\rho_1 + \rho_2 - 1)^2}{\rho_1 (\rho_1 + \rho_2) (\rho_2 - 1)}.
\end{align*}}

\begin{proposition}[Task model distance]\label{prop_dist_transition}
	In the isotropic model, suppose that $\rho_1 > 1$ and $\sigma_1 = \sigma_2 = \sigma$.
	Then we have that
	%Whether $\te(\hat{\beta}_t^{\MTL})$ is lower than $\te(\hat{\beta}_t^{\STL})$ is determined by the ratio between $\Psi(\beta_1, \beta_2)$ and $\Phi(\rho_1, \rho_2)$:
	\squishlist
		\item If $\Psi(\beta_1, \beta_2) < \nu \cdot  \Phi(\rho_1, \rho_2)$, then w.h.p. over the randomness of $X_1,X_2$, $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\Psi(\beta_1, \beta_2) > \frac{1}{\nu} \cdot  \Phi(\rho_1, \rho_2)$, then w.h.p. over the randomness of $X_1,X_2$, $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\squishend
	Here $\nu = (1-\oo(1))\min(\cdot(1 - \rho_1^{-1/2})^4, (1 + \rho_1^{-1/2})^{-4})$.
	Concretely, if $\rho_1 > 40$, then $\nu\in (1,2)$.
\end{proposition}

Proposition \ref{prop_dist_transition} simplifies Theorem \ref{thm_main_informal} in the isotropic model, allowing for a more explicit statement of the bias-variance tradeoff.
Concretely, $\Psi(\beta_1, \beta)$ and $\Phi(\rho_1, \rho_2)$ corresponds to $\Delta_{\bias}$ and $\Delta_{\vari}$, respectively.
Prosposition \ref{prop_dist_transition} explains the transition observed in Figure \ref{fig_model_shift}.
Note that there are several unexplained observations near the transition threshold $0$, which are caused by the concentration error $\nu$.
%We fix the target task and vary the source task, in particular the parameter $d$ which determines $\norm{\beta_1 - \beta_2}$.
%Figure \ref{fig_model_shift} shows the result.
%We observe that Proposition \ref{prop_dist_transition} explains most of the observations in Figure \ref{fig_model_shift}.
%The proof of Proposition \ref{prop_dist_transition} involves two parts.
%First, in equation \eqref{eq_te_var}, the positive variance reduction effect scales with $n_1 = \rho_1 p$, the number of source task data points.
%Second, we show that the negative effect of model-shift bias scales with $pd^2$, which is the expectation of $\norm{\beta_1 - \beta_2}^2$.
The proof of Proposition \ref{prop_dist_transition} can be found in Appendix \ref{app_proof_31}. %, which is based on our main result described later in Theorem \ref{thm_model_shift}.

@@

In classical Rademacher or VC based theory of multi-task learning, the generalization bounds are usually presented for settings where the sample sizes are equal for all tasks \cite{B00,M06,MPR16}.
More generally, such results are still applicable when all task data are being added simultaneously.
On the other hand, for many applications of multi-task learning, the data sources are usually heterogeneous.
For such settings, we have observed that adding more labeled data from one task does not always help.
%On the other hand, we have observed that adding more labeled data does not always improve performance in multi-task learning.
Using the isotropic model, we consider what happens if we vary the source task sample size.
%Figure \ref{fig_size} provides a simulation result for such a setting.
%We observe that as $n_1 / n_2$ increases, there is a transition from positive to negative transfer.
Our result below explains the phenomenon of Figure \ref{fig_size}.

\begin{proposition}[Source/target sample ratio]\label{prop_data_size}
	In the isotropic model, suppose that $\rho_1 > 40$ and $\rho_2 > 110$ are fixed constants, and $\Psi(\beta_1, \beta_2) > 2/(\rho_2 - 1)$.
	Then we have that
	\squishlist
		\item If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} < {\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 1) - \nu^{-1}}$, then w.h.p. $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} > \frac{1}{\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 1.5) - \nu^{-1}}$, then w.h.p. $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\squishend
\end{proposition}
Proposition \ref{prop_data_size} describes the bias-variance tradeoff in terms of the data ratio $n_1 / n_2$.
From Figure \ref{fig_size}, we see that our result is able to explain the transition from positive to negative transfer.
There are several unexplained observations near $y = 0$ caused by the errors $\nu$.
The proof of Proposition \ref{prop_data_size} can be found in Appendix \ref{app_proof_32}.

\textbf{Connection to Taskonomy.} We use our tools to explain a key result of Taskonomy by Zamir et al. (2018) \cite{ZSSGM18}, which shows that MTL can reduce the amount of labeled data needed to achieve comparable performance to STL.
%More precisely, suppose we have $n_i$ datapoints for each task, for $i= 1, 2$.
For $i = 1, 2$, let $\hat{\beta}_i^{\MTL}(x)$ denote the estimator trained using $x \cdot n_i$ datapoints from every task. The data efficiency ratio is defined as
{\small\[ \argmin_{x\in(0, 1)} ~~
		\te_1(\hat{\beta}_1^{\MTL}(x)) + \te_2(\hat{\beta}_2^{\MTL}(x))
		\le \te_1(\hat{\beta}_1^{\STL}) + \te_2(\hat{\beta}_2^{\STL}). \]}
For example, the data efficiency ratio is $1$ if there is negative transfer.
Using our tools, we show that in the isotropic model, the data efficiency ratio is
roughly
{\small\[ \frac{1}{\rho_1 + \rho_2} {+ \frac{2}{(\rho_1 +\rho_2)(\rho_1^{-1} + \rho_2^{-1} - \Theta(\Psi(\beta_1, \beta_2)))}}. \]}
Compared with Proposition \ref{prop_dist_transition}, we see that when $\Psi(\beta_1, \beta_2)$ is smaller than $\rho_1^{-1} + \rho_2^{-1}$ (up to a constant multiple), the transfer is positive.
Moreover, the data efficiency ratio quantifies how effective the positive transfer is using MTL.
%In addition to $\rho_1,\rho_2$, the data efficiency ratio also scales with $\Psi(\beta_1, \beta_2)$, the model distance between the tasks.
The result can be found in Proposition \ref{prop_data_efficiency} in Appendix \ref{app_proof_32}.

>>>>

\textit{The isotropic model.}
	Consider two tasks with isotropic covariances $\Sigma_1 = \Sigma_2 = \id$.
	Each task has sample size $n_i = \rho_i \cdot p$, for $i = 1, 2$.
	%And $X_1\in\real^{n_1\times p}, X_2\in\real^{n_2\times p}$ denote the covariates of the two tasks, respectively.
	We assume that for the target task, $\beta_2$ has i.i.d. entries with mean zero and variance $\kappa^2$.
	For the source task, $\beta_1 $ equals $\beta_2$ plus i.i.d. entries with mean $0$ and variance $d^2$.
	The labels are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\e_i$ consists of i.i.d. entries with mean zero and variance $\sigma^2$.
%	We assume that all the random variables have subexponential decay, while keeping in mind that our results can be applied under weaker moments assumptions as shown in Appendix \ref{sec_maintools}.

In the isotropic model, we show that as we increase the distance between $\beta_1$ and $\beta_2$ (or $d$), there is a transition from positive transfer to negative transfer in MTL.
%We measure model dissimilarity as $\norm{\beta_1 - \beta_2}^2$, which is the distance between source and target in the isotropic model.
%Figure \ref{fig_model_shift} provides a simulation when $p = 200$.
%{The rest of parameter settings can be found in Appendix \ref{app_synthetic}.}
%Our result below will provide an explanation to this phenomenon.
%Based on Theorem \ref{thm_model_shift}, we derive the transition threshold in the following proposition.
We introduce the following notations.
{\small\begin{align*}
	\Psi(\beta_1, \beta_2) = {\ex{\bignorm{\beta_1 - \beta_2}^2}} / {\sigma^2},  \quad \Phi(\rho_1, \rho_2) = \frac{(\rho_1 + \rho_2 - 1)^2}{\rho_1 (\rho_1 + \rho_2) (\rho_2 - 1)}.
\end{align*}}

\begin{proposition}[Task model distance]\label{prop_dist_transition}
	In the isotropic model, suppose that $\rho_1$ and $\rho_2 > 1$.
	Then
	%Whether $\te(\hat{\beta}_t^{\MTL})$ is lower than $\te(\hat{\beta}_t^{\STL})$ is determined by the ratio between $\Psi(\beta_1, \beta_2)$ and $\Phi(\rho_1, \rho_2)$:
	\squishlist
		\item If $\Psi(\beta_1, \beta_2) < \nu \cdot  \Phi(\rho_1, \rho_2)$, then w.h.p. over the randomness of $X_1,X_2$, $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\Psi(\beta_1, \beta_2) > \frac{1}{\nu} \cdot  \Phi(\rho_1, \rho_2)$, then w.h.p. over the randomness of $X_1,X_2$, $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\squishend
	Here $\nu = (1-\oo(1))\min(\cdot(1 - \rho_1^{-1/2})^4, (1 + \rho_1^{-1/2})^{-4})$.
	Concretely, if $\rho_1 > 40$, then $\nu\in (1,2)$.
\end{proposition}

Proposition \ref{prop_dist_transition} simplifies Theorem \ref{thm_main_informal} in the isotropic model, allowing for a more explicit statement of the bias-variance tradeoff.
Concretely, $\Psi(\beta_1, \beta)$ and $\Phi(\rho_1, \rho_2)$ corresponds to $\Delta_{\bias}$ and $\Delta_{\vari}$, respectively.
We apply Proposition \ref{prop_dist_transition} to the parameter setting of Figure \ref{fig_model_shift} (the details are left to Appendix \ref{app_synthetic}).
We can see that our result is able to predict positive or negative transfer  accuractly that matches the empirical curve.
There are several unexplained observations near the transition threshold $0$, which are caused by the concentration error $\nu$.
%We fix the target task and vary the source task, in particular the parameter $d$ which determines $\norm{\beta_1 - \beta_2}$.
%Figure \ref{fig_model_shift} shows the result.
%We observe that Proposition \ref{prop_dist_transition} explains most of the observations in Figure \ref{fig_model_shift}.
%The proof of Proposition \ref{prop_dist_transition} involves two parts.
%First, in equation \eqref{eq_te_var}, the positive variance reduction effect scales with $n_1 = \rho_1 p$, the number of source task data points.
%Second, we show that the negative effect of model-shift bias scales with $pd^2$, which is the expectation of $\norm{\beta_1 - \beta_2}^2$.
The proof of Proposition \ref{prop_dist_transition} can be found in Appendix \ref{app_proof_31}. %, which is based on our main result described later in Theorem \ref{thm_model_shift}.

@@

In classical Rademacher or VC based theory of multi-task learning, the generalization bounds are usually presented for settings where the sample sizes are equal for all tasks \cite{B00,M06,MPR16}.
%More generally, such results are still applicable when all task data are being added simultaneously.
On the other hand, uneven sample sizes between different tasks (or even dominating tasks) have been empirically observed as a cause of negative transfer \cite{YKGLHF20}.
For such settings, we have also observed that adding more labeled data from one task does not always help.
%On the other hand, we have observed that adding more labeled data does not always improve performance in multi-task learning.
Using the isotropic model, we consider what happens if we vary the source task sample size.
%Figure \ref{fig_size} provides a simulation result for such a setting.
%We observe that as $n_1 / n_2$ increases, there is a transition from positive to negative transfer.

\begin{proposition}[Source/target sample ratio]\label{prop_data_size}
	In the isotropic model, suppose that $\rho_1 > 40$ and $\rho_2 > 110$ are fixed constants, and $\Psi(\beta_1, \beta_2) > 2/(\rho_2 - 1)$.
	Then we have that
	\squishlist
		\item If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} < {\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 1) - \nu^{-1}}$, then w.h.p. $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} > \frac{1}{\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 1.5) - \nu^{-1}}$, then w.h.p. $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\squishend
\end{proposition}
Proposition \ref{prop_data_size} describes the bias-variance tradeoff in terms of the data ratio $n_1 / n_2$.
We apply the above result to the parameter setting of Figure \ref{fig_size} (described in Appendix \ref{app_synthetic}).
Our theory accuractly predicts a curious phenomenon, where increasing the sample size of the source task results in negative transfer!
There are several unexplained observations near $y = 0$ caused by the errors $\nu$.
The proof of Proposition \ref{prop_data_size} can be found in Appendix \ref{app_proof_32}.

\textbf{Connection to Taskonomy.} We use our tools to explain a key result of Taskonomy by Zamir et al. (2018) \cite{ZSSGM18}, which shows that MTL can reduce the amount of labeled data needed to achieve comparable performance to STL.
%More precisely, suppose we have $n_i$ datapoints for each task, for $i= 1, 2$.
For $i = 1, 2$, let $\hat{\beta}_i^{\MTL}(x)$ denote the estimator trained using $x \cdot n_i$ datapoints from every task. The data efficiency ratio is defined as
{\small\[ \argmin_{x\in(0, 1)} ~~
		\te_1(\hat{\beta}_1^{\MTL}(x)) + \te_2(\hat{\beta}_2^{\MTL}(x))
		\le \te_1(\hat{\beta}_1^{\STL}) + \te_2(\hat{\beta}_2^{\STL}). \]}
For example, the data efficiency ratio is $1$ if there is negative transfer.
Using our tools, we show that in the isotropic model, the data efficiency ratio is
roughly
{\small\[ \frac{1}{\rho_1 + \rho_2} {+ \frac{2}{(\rho_1 +\rho_2)(\rho_1^{-1} + \rho_2^{-1} - \Theta(\Psi(\beta_1, \beta_2)))}}. \]}%
Compared with Proposition \ref{prop_dist_transition}, we see that when $\Psi(\beta_1, \beta_2)$ is smaller than $\rho_1^{-1} + \rho_2^{-1}$ (up to a constant multiple), the transfer is positive.
Moreover, the data efficiency ratio quantifies how effective the positive transfer is using MTL.
%In addition to $\rho_1,\rho_2$, the data efficiency ratio also scales with $\Psi(\beta_1, \beta_2)$, the model distance between the tasks.
The result can be found in Proposition \ref{prop_data_efficiency} in Appendix \ref{app_proof_32}.
